\textsc{Atomic modeling of argon}\\ \footnotesize{\textsc{Project 5 --- FYS3150}}

### 2015-12-04 projects/project_5/molecular_dynamics/article/article.tex
fixed units on temperature ratio

\documentclass[a4paper]{article}
\usepackage[]{enumerate} 
\usepackage[]{graphicx} 
\usepackage[]{subcaption} 
\usepackage[]{amsthm, amsmath, amssymb} 
\usepackage[]{fancyhdr}
\usepackage[]{hyperref} 
\urlstyle{sf}
\graphicspath{{../pictures/}}

@@

\begin{abstract}
    In this project we explore the world of molecular dynamics. We are first
    presented with the skeleton of a working MD-program, and we are asked to
    implement the functionality we need as we go along. We look at the time
    evolution of a system of argon, with an initial face-centered cubic lattice
    configuration. The atoms are under the influence of the Lennard-Jones
    potential. We use the Velocity-Verlet as our time integrator of choice, and
    we compare this to the Euler-Cromer method in terms of energy conservation
    properties. We also compute the melting temperature of argon, by means of
    the Einstein relation and diffusion constant.
\end{abstract}
\tableofcontents
\listoffigures

@@

    The problem that arises when using the Maxwell-Boltzmann distribution is
    that in its initial state the system has a non-zero net momentum. This
    causes the system to drift. This is why we want to implement the function
    \texttt{removeMomentum} in the \texttt{System}-class. Its purpose is to
    compute the total momentum of the system, and then remove a small portion
    of momentum from each atom in order to have a net momentum of zero.

    We compute the velocity of the center of mass of the system, and then
    subtract this velocity from each atom.  Let $m_i$ and $v_i$ denote the mass
    and velocity of atom $i$ respectively and let $M$ and $V$ denote the total
    mass of the system and the velocity of the center of mass. Then
    \begin{align*}
        \notag
        V = \sum^{n}_{i=1} m_i v_i, \quad M = \sum^{n}_{i=1} m_i
    \end{align*}
    where $n$ is the total number of atoms in the system. We then need to
    subtract the value $V_{\mathrm{sub}} = V / M$ from each $v_i$.

@@

    We divide the domain of our system into \emph{cells}, each with an edge
    length of at least the cut-of radius of the potential. This ensures that if
    we only compute the mutual forces between atoms in the same or adjacent
    cells, then the atoms are never a distance farther than $r_\mathrm{cut}$
    apart. For each integration step we check the validity of the list of
    neighbors for each atom by iterating through each atom in the system, and
    checking whether it has moved a distance larger than $r_\mathrm{cut}$ since
    the last time the list of neighbors was built.

@@

    Argon has a melting point of $\approx 84$K and a boiling point of $\approx
    87$K. If we run the simulations, once for a temperature of $40$K, once for
    $85$K and finally once for $120K$ we should see indications of three
    different phases. The results shows that the model somewhat accurately
    manages to capture these phase transitions. Snapshots taken at the same
    point in time for all three cases are presented in figure
    \ref{fig:potential_iteration}.
    \begin{figure}
        \begin{tabular}[c]{ccc}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{solid_40K.png}
                \caption{$T = 40$K}
            \end{subfigure}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{liquid_85K.png}
                \caption{$T = 85$K}
            \end{subfigure}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{gas_120K.png}
                \caption{$T = 120$K}
            \end{subfigure}
        \end{tabular}
        \caption[Argon phases]{Snapshots taken of the system at frame 500/999 for
            temperatures of $40$, $85$ and $120$ kelvin. It is a bit difficult
            to conclude from the pictures, one really needs to see the system
            in motion, but the system do behave more or less like a solid, a
            liquid and a gas for respective temperatures. It is possible to see a
            certain change in the entropy of the system across the three
            images. These made use of the Velocity-Verlet integration method.}
        \label{fig:potential_iteration}
    \end{figure}

@@

    \begin{figure}[h]
        \centering \includegraphics[width=\linewidth]{temperature_ratio.pdf}
        \caption[Temperature ratio]{Plotting the ratio between initial and
            instantaneous temperature for varying $T_i$. Here the initial
            temperatures ranges from 20 to 500 kelvin, with 48 intermediate steps.
        As we can see the ratio stabilizes around 0.5. The graph of an arbitrary
    temperature has been highlighted in black.}
        \label{fig:temperature_ratio}
    \end{figure}

@@

    \begin{figure}[h]
        \centering
        \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
        \caption[Diffusion constant]{The diffusion constant as a function of temperature. We here
                see that $D$ abruptly increases at the temperature of $T = 304K$
                which indicates that the crystal is melting.  The two lines are the
                least square solutions to the noisy data before and after,
                respectively. The diffusion constant is given in units of area per second.}
        \label{fig:diffusion_constant}
    \end{figure}
\clearpage

>>>>

\documentclass[a4paper]{article}
\usepackage[]{enumerate} 
\usepackage[]{graphicx} 
\usepackage[]{subcaption} 
\usepackage[]{amsthm, amsmath, amssymb} 
\usepackage[]{fancyhdr}
\usepackage[]{setspace} 
\usepackage[]{hyperref} 
\urlstyle{sf}
\graphicspath{{../pictures/}}

@@

\begin{abstract}
    In this project we explore the world of molecular dynamics. We are first
    presented with the skeleton of a working MD-program, and we are asked to
    implement the functionality we need as we go along. We look at the time
    evolution of a system of argon, with an initial face-centered cubic lattice
    configuration. The atoms are under the influence of the Lennard-Jones
    potential. We use the Velocity-Verlet as our time integrator of choice, and
    we compare this to the Euler-Cromer method in terms of energy conservation
    properties. We also compute the melting temperature of argon, by means of
    the Einstein relation and diffusion constant.
\end{abstract}
\doublespacing
\tableofcontents
\listoffigures      
\singlespacing

@@

    As shown in figure \ref{fig:maxwell_boltzmann}, increasing the temperature
    yields a more even velocity distribution. 
    The problem that arises when using the Maxwell-Boltzmann distribution is
    that in its initial state the system has a non-zero net momentum. This
    causes the system to drift. This is why we want to implement the function
    \texttt{removeMomentum} in the \texttt{System}-class. Its purpose is to
    compute the total momentum of the system, and then remove a small portion
    of momentum from each atom in order to have a net momentum of zero.

    \begin{figure}[h]
        \centering \includegraphics[width=0.7\linewidth]{maxwell_boltzmann.pdf}
        \caption[The Maxwell-Boltzmann distribution]{The Maxwell-Boltzmann
            distribution for a varying scale parameter $a = k_BT/m$.  As we can
            see, increasing the temperature increases the likelihood of an atom
        being assigned an arbitrary velocity, where as for low temperatures the
    distribution is biased towards one specific velocity.}
        \label{fig:maxwell_boltzmann}
    \end{figure}
    We compute the velocity of the center of mass of the system, and then
    subtract this velocity from each atom.  Let $m_i$ and $v_i$ denote the mass
    and velocity of atom $i$ respectively and let $M$ and $V$ denote the total
    mass of the system and the velocity of the center of mass. Then
    \begin{align*}
        \notag
        V = \sum^{n}_{i=1} m_i v_i, \quad M = \sum^{n}_{i=1} m_i
    \end{align*}
    where $n$ is the total number of atoms in the system. We then need to
    subtract the value $V_{\mathrm{sub}} = V / M$ from each $v_i$.

@@

    We divide the domain of our system into \emph{cells}, each with an edge
    length of at least the cut-of radius of the potential. This ensures that if
    we only compute the mutual forces between atoms in the same or adjacent
    cells, then the atoms are never a distance farther than $r_\mathrm{cut}$
    apart. For each integration step we check the validity of the list of
    neighbors for each atom by iterating through each atom in the system, and
    checking whether it has moved a distance larger than $r_\mathrm{cut}$ since
    the last time the list of neighbors was built. It is interesting to note
    that we have to rebuild the neighbor lists more often for higher
    temperature due to the velocities being proportional to the temperature.

@@

    Argon has a melting point of $\approx 84$K and a boiling point of $\approx
    87$K. If we run the simulations, once for a temperature of $80$K, once for
    $170$K and finally once for $240K$ we should see indications of three
    different phases (see section \ref{sub:initial_to_final_temperature} for
    the choices of these temperatures). The results shows that the model does
    not manage to capture these phase transitions. This is probably attributed
    to the Lennard-Jones potential. If we run the simulations for initial
    temperatures of $400K, 700K$ and $900K$ we should see indications of phase
    change. We will get back to this in section
    \ref{sub:melting_temperature_of_system}. Snapshots taken at the same point
    in time for all three cases are presented in figure
    \ref{fig:potential_iteration}.
    \begin{figure}
        \begin{tabular}[c]{ccc}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{solid_40K.png}
                \caption{$T = 400$K}
            \end{subfigure}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{liquid_85K.png}
                \caption{$T = 700$K}
            \end{subfigure}
            \begin{subfigure}[c]{0.32\textwidth}
                \includegraphics[height=100pt, width=\linewidth]{gas_120K.png}
                \caption{$T = 900$K}
            \end{subfigure}
        \end{tabular}
        \caption[Argon phases]{Snapshots taken of the system at frame 500/999
            for temperatures of $400$, $700$ and $900$ kelvin. It is a bit
            difficult to conclude from the pictures, one really needs to see
            the system in motion, but the system do behave more or less like a
            solid, a liquid and a gas for respective temperatures. It is
        possible to see a certain change in the entropy of the system across
    the three images. These made use of the Velocity-Verlet integration
method.}
        \label{fig:potential_iteration}
    \end{figure}

@@

    \begin{figure}[h]
        \centering \includegraphics[width=\linewidth]{temperature_ratio.pdf}
        \caption[Temperature ratio]{Plotting the ratio between initial and
            instantaneous temperature for varying $T_i$. Each individual graph
            corresponds to one value of $T_i$ and the ratio is plotted against
            time given in units of tenths of a picosecond. Here the initial
            temperatures ranges from 20 to 500 kelvin, with 48 intermediate
            steps.  As we can see the ratio stabilizes around $1/2$. The graph of
            an arbitrary temperature has been highlighted in black.}
        \label{fig:temperature_ratio}
    \end{figure}

@@

    \begin{figure}[h]
        \centering
        \includegraphics[width=0.8\linewidth]{diffusion_constant.pdf}
        \caption[Diffusion constant]{The diffusion constant as a function of
            temperature. We here see that $D$ abruptly increases at the
            temperature of $T = 304K$ which indicates that the crystal is
        melting. The two lines are the linear graphs best fitted to the noisy
    data.  The diffusion constant is given in units of area per second.}
        \label{fig:diffusion_constant}
    \end{figure}
\clearpage
